Supplement to ``{\mytitle}''

### 2023-09-05 si/si.Rnw

\pdfminorversion=4

\newcommand\secTitleSpace{\hspace{3mm}}

<<packages, include=FALSE>>=
library(tidyverse)
library(pomp)
library(haitipkg)
library(spatPomp)
library(doParallel)
library(doRNG)

@@

<<set-opts, include=FALSE>>=
options(
  scipen = 2,
  help_type = "html",
  stringsAsFactors = FALSE,
  continue = "+  ",
  width = 70,
  useFancyQuotes = FALSE,
  reindent.spaces = 2,
  xtable.comment = FALSE
)
@

<<knitr-opts, include=FALSE, purl=FALSE>>=
library(knitr)
opts_knit$set(concordance=TRUE)
opts_chunk$set(
  progress=TRUE,prompt=TRUE,highlight=FALSE,
  tidy=TRUE,
  tidy.opts=list(
    keep.blank.line=FALSE
  ),
  comment="",
  warning=FALSE,
  message=FALSE,
  error=TRUE,
  echo=FALSE,
  cache=FALSE,
  strip.white=TRUE,
  results="markup",
  fig.path="figure/",
  fig.lp="fig:",
  fig.align="center",
  fig.show="asis",
  dev="pdf",
  dev.args=list(
    bg="transparent",
    pointsize=9
  )
)

myround <- function (x, digits = 1) {
  # taken from the broman package
  if (digits < 1)
    stop("This is intended for the case digits >= 1.")
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
  tmp[tmp == paste0("-", zero)] <- zero
  tmp
}

@

%% copied from pnas-new.cls
\usepackage{amsmath,amsfonts,amssymb,graphicx}

%%%% parameters %%%%%%%%%%5
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\Whur[1]{\beta_{W#1}^{hm}}
\newcommand\hHur[1]{h_{#1}^{hm}}
\newcommand\Iinit{I_{0,0}}
\newcommand\Einit{E_{0, 0}}
\newcommand\reportChange{\delta}

\newcommand\missing{\hspace{6mm} ---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{\bullet}

\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}

%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5

\newcommand\siOnly[1]{} %% redefined for the supplement
\newcommand\msOnly[1]{#1} %% redefined for the supplement

\newcommand\code[1]{\texttt{#1}}
\usepackage{xurl}
% \usepackage[breaklinks]{hyperref}
\usepackage{makecell}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[ruled,noline,linesnumbered]{algorithm2e}
% \usepackage{setspace}

%% EDITING MACROS
\usepackage{color}
\usepackage[normalem]{ulem}% to use \sout in feedback commands
\usepackage{bm}
% orange for EI
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
\newcommand\eic[1]{\textcolor{orange}{[#1]}}
% green for JW
\definecolor{green}{rgb}{0,0.5,0}
\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
\newcommand\jwc[1]{\textcolor{green}{[#1]}}
% purple for JJ
\definecolor{purple}{rgb}{0.5,0,1}
\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
% cyan for AR
\definecolor{cyan}{rgb}{0,.5,.5}
\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
\newcommand\arc[1]{\textcolor{cyan}{#1}}
% light brown for KT
\definecolor{lightbrown}{rgb}{0.5,0.5,0}
\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}

%% basic POMP definitions
\newcommand\Xspace{{\mathbb X}}
\newcommand\Yspace{{\mathbb Y}}
\newcommand\Thetaspace{\R^{\Thetadim}}
\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
\newcommand\hatTheta{\widehat{\Theta}}
\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
\newcommand\Thetadim{P}
\newcommand\thetadim{p}
\newcommand\timeSet{{\mathbb T}}
\newcommand\data[1]{#1^*}

\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}

% for algorithm pseudocode
\SetKwFor{For}{for}{:}{end}

%% for particle filters
\newcommand\unit{u}
\newcommand\altUnit{\tilde u}
\newcommand\Unit{U}
\newcommand\rep{i}
\newcommand\Rep{\mathcal{I}}
\newcommand\island{\rep}
\newcommand\Island{\Rep}
%\newcommand\txtisland{replicate}
\newcommand\altIsland{\tilde i}
\usepackage[mathscr]{euscript}

\renewcommand\time{n}
\newcommand\myvec[1]{\boldsymbol{#1}}
\newcommand\altTime{\tilde n}
\newcommand\Time{N}
\newcommand\Np{J}
\newcommand\np{j}
\newcommand\altNp{k}
\newcommand\altAltNp{\tilde j}
\newcommand\resampleIndex{r}
\newcommand\unitWeight{w}

%% for all iterated filtering algorithms
\newcommand\Nit{M}
\newcommand\nit{m}

%% customized math macros
\newcommand\prob{\mathbb{P}}
\newcommand\E{\mathbb{E}}
\newcommand\dd[1]{\mathrm{d}{#1}}
\newcommand\given{{\,\vert\,}}
\newcommand\equals{{{\,}={\,}}}
\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
\newcommand\myto{{\;:\;}}
\newcommand\seq[2]{{#1}\!:\!{#2}}
\newcommand\mydot{{\,\cdot\,}}
\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
\newcommand\normal{\mathcal{N}}
\newcommand\argequals{{\,=\,}}
\newcommand\lags{c}
\newcommand\maxlag{\overline{c}}
\newcommand\nlfList{C}
\newcommand\bigO{\mathcal{O}}
\newcommand\loglik{\lambda}
\newcommand\loglikMC{\MC{\loglik}}
\newcommand\R{\mathbb{R}}
\newcommand\param{\,;}
\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
\newcommand\EMC{{\E}}
\newcommand\Var{\mathrm{Var}}
\newcommand\var{\Var}
\newcommand\Cov{\mathrm{Cov}}
\newcommand\cov{\Cov}
\newcommand\iid{\mathrm{iid}}
\newcommand\dist{d}
\newcommand\transpose{\top}

\newcommand\gnsep{,}

\def\lik{L}
\def\loglik{\ell}

%% THEOREM-LIKE ENVIRONMENTS
\usepackage{amsthm}
\newtheorem{prop}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{example}{Example}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corrolary}
\newtheorem{fact}{Fact}
\newtheorem{assumption}{Assumption}
\newtheorem{assumptionB}{Assumption}
\renewcommand\theassumption{A\arabic{assumption}}
\renewcommand\theassumptionB{B\arabic{assumptionB}}

%% FOR PSEUDOCODE SETUP
\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
\newcommand\asp{\hspace{4mm}}
\newcommand\codeIndent{\hspace{4mm}}

%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%

\bibliographystyle{apalike}
\usepackage{natbib}
\usepackage{fullpage}

\newcommand\myeqref[1]{(\ref{#1})}

\renewcommand{\contentsname}{Supplementary Content}
\renewcommand{\refname}{Supplementary References}
\renewcommand\thefigure{S-\arabic{figure}}
\renewcommand\thetable{S-\arabic{table}}
\renewcommand\thesection{S\arabic{section}}
\renewcommand\theequation{S\arabic{equation}}
\renewcommand\theprop{S\arabic{prop}}
\renewcommand\thelemma{S\arabic{lemma}}
\renewcommand\thealgocf{S\arabic{algocf}}

@@

\section{Markov chain and differential equation interpretations of compartment flow rates}

In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
For a discrete population model, these rates define a Markov chain.
For a continuous and deterministic model, the rates define a system of ordinary differential equations.
Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
Our treatment follows \citet{breto09}.

A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
These flows are required to satisfy a ``conservation of mass'' identity:
\begin{equation}
X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
\end{equation}
Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.

There are different ways to use a collection of rate functions to build a fully specified model.
We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}). Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.

\subsection{Ordinary differential equation (ODE) interpretation}
\label{subsec:ode}

A basic deterministic specification is
\begin{equation}
\label{eq:ode1}
dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
\end{equation}
where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
Instead, we specify
\begin{equation}
\label{eq:ode2}
dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
\end{equation}
This is the the interpretation and implementation used for Model~2 in our study.

\subsection{Simple Markov counting system interpretation}
\label{subsec:smcs}
A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
A basic approach to this is to define
\begin{eqnarray}
\label{eq:smcs1}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
\\
\label{eq:smcs2}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
 &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
\end{eqnarray}
for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
As with the ODE case, we need special attention for flows into the system, and we define
\begin{eqnarray}
\label{eq:smcs3}
\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs4}
\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
 &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
\end{eqnarray}
Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
We therefore call the Markov chain a simple Markov counting system (SMCS).
The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.

\subsection{Overdispersed Markov counting system interpretation}
\label{subsec:odmcs}

Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
Specifically, we consider a model given by
\begin{equation}
\label{eq:odmcs1}
\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
\end{equation}
where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
\begin{equation}
\label{eq:odmcs2}
\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
\end{equation}
Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.

\section{Numerical solutions to compartment models}
\label{sec:numerics}

Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
However, for applicability to larger populations, we use instead the following Euler scheme.
Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
\begin{equation}
\label{eq:numerics1}
\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
\end{equation}
In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
Then, for $i\neq \demography$ and $j\neq i$, and writing
\begin{equation}
\label{eq:numerics2}
\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
\end{equation}
we calculate transition probabilities
\begin{eqnarray}
\label{eq:numerics3}
p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
\\
\label{eq:numerics4}
p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
\end{eqnarray}
These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.

Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
For the case $i=\demography$, the treatment used by Model~1 is to set
\begin{equation}
\label{eq:numerics6}
\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
\end{equation}
an independent Poisson random variable with mean $\mu_{\demography j} \delta$.

Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.

\section{Measurement Models}

Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.

@@

In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
\begin{equation}
  \label{model1:meas}
  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
\end{equation}
where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.

\subsection{Model~2}\label{sec:mod2Meas}

Model~2 was fit using reported case counts that were transformed using the natural logarithm.
We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
\begin{equation}
  \label{model2:meas}
  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
\end{equation}
where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.

This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
The measurement model they used for the epidemic phase is
\begin{equation}
  \label{model2:measEpi}
  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
\end{equation}
% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
\begin{equation}
  \label{model2:measEnd}
  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
\end{equation}
where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.

@@

In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
\begin{equation}
  \label{model3:meas}
  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
\end{equation}
where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.

This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
We therefore do not allow a change in reporting rate when fitting Model~3.

\section{Initial Values}

To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
This density is used to obtain initial values of the latent state when fitting and evaluating the model.
For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
In the following subsections, we mention initial value parameters that were fit for each model.

For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
This model for the initial values of the latent states matches that which was used by \citet{lee20}.

Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
This model for the initial values of the latent states matches that which was used by \citet{lee20}.

The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.

We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.

\begin{eqnarray}
\label{eq:model3:I0}
I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
\\%[-3pt]
\label{eq:model3:A0}
A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
\\
\label{eq:model3:R0}
R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
\\
\label{eq:model3:S0}
S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
\\
\label{eq:model3:B0}
B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
\end{eqnarray}

In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.

% Argue why we didn't need to do this for Model 2.

% TODO: Check this section.

\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}

<<CalibrateOrig3>>=
# Create vectors for the unit and shared parameters
unit_specific_names <- c("betaB", "foi_add")
shared_param_names <- c(
  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
  "epsilon", "k"
)
est_param_names <- c(
  unit_specific_names, shared_param_names
)

# Add unit numbers to each parameter
est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)

# Simple function that is used to set the rw_sd for the first search
# This function is convenient so that we don't have to type out the rw_sd
# for each unit parameter.
set_rw_1 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.01)
  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
  # } else if (x %in% c("Iinit3", "Iinit4")) {
  #   return(expression(ivp(0.25)))
  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
    return(0.015)
  } else if (grepl('^betaB[[:digit:]]+$', x)) {
    return(0.01)
  } else {
    return(0)
  }
}

set_rw_2 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.003)
  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
  # } else if (x %in% c("Iinit3", "Iinit4")) {
  #   return(expression(ivp(0.125)))
  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
    return(0.005)
  } else if (grepl('^betaB[[:digit:]]+$', x)) {
    return(0.003)
  } else {
    return(0)
  }
}

>>>>

\input{inputs/header}

<<packages, include=FALSE>>=
library(tidyverse)
library(pomp)
library(haitipkg)
library(spatPomp)
library(doParallel)
library(doRNG)
library(knitr)

options(
  scipen = 2,
  help_type = "html",
  stringsAsFactors = FALSE,
  continue = "+  ",
  width = 70,
  useFancyQuotes = FALSE,
  reindent.spaces = 2,
  xtable.comment = FALSE
)

opts_knit$set(concordance=TRUE)
opts_chunk$set(
  tidy.opts=list(
    keep.blank.line=FALSE
  ),
  comment="",
  # warning=FALSE,
  # message=FALSE,
  echo=FALSE,
  fig.align="center",
  dev="pdf",
  dev.args=list(
    bg="transparent",
    pointsize=9
  )
)

@@

@@

\section{Model Diagrams}

Each of the dynamic models considered in this manuscript can be fully described using the model descriptions in the manuscript, coupled with the additional information described in Sections 2 and 3 of this supplement.
Despite this, diagrams of dynamic systems are often helpful to understand the equations.
In this section, we give three diagrams representing Models~1--3, respectively.
Because the models are defined by their mathematical equations and numeric implementation, these diagrams are not unique visual representations of the model.
Alternative representations that may be helpful in understanding the models explored in this paper were made by \citet{lee20sup}.

@@

\input{inputs/mod1diagram}

\input{inputs/mod2diagram}

@@

\input{inputs/mod3diagram}

\input{inputs/modelDetails}

\input{inputs/measurementModels}

\input{inputs/initialValues}

\input{calibrateMod3Out}
